A Distributed Segment-based Approach for Topology-agnostic Routing in Self-Assembled Nano Networks

### 2013-05-13 trunk/PAPER/paper.tex
- completed intro draft

\title{A Distribuited Segment-based Approach for Topology-agnostic Nano Networks Routing}

\author{
  \IEEEauthorblockN{a} \IEEEauthorblockA{Affiliation3\\ email3} \and
  \IEEEauthorblockN{b} \IEEEauthorblockA{Affiliation2\\ email2} }

@@

\begin{abstract}
In this paper we present a first effort for a distributed
segment-based approach to routing which exploits the already-proven
qualities of segment-based routing without requiring as graph topology
as input, and without the need for a centralized algorithm which
perform a setup phase.  Our contribution is not find the
best segment choice, but to show something that could really work in
such complex, irregular and large sized topologies.
\end{abstract}

\IEEEkeywords{...}

\section{Introduction}
Exploring long term alternatives to the CMOS technology is gaining
more and more revelance as the scaling trend of such devices continue
to introduce new challenges. Power density, defect tolerance, testing
costs and many other issues are only partially mitigated by multicore
approaches~\ref{TODO}, but growing computing requirements will eventually need
even more radical architectural modifications and new paradigms in
future computer design.

DNA-self assembled nanoscale architectures~\ref{TOOD} are emerging as promising
technology due its tera/peta scale of integration, defect tolerance
tolerance and potential computing capabilities~\ref{TODO}. This
technologies is certainly still at its early stage of development,
however different laboratory demos and prof-of-concepts architectures
have been presented~\ref{TODO}.

The main idea beyond this approach is to exploit the regularity and
stabiliy of DNA sequences in orders to create grids onto which
nano-devices (e.g. nanowires and CNFET~\ref{TODO}) can be attached by
designing approapriate DNA tags. A detailed description of the
chemical properties involved is beyond the scope of this paper, since
we focus of the main challenges that this new fabrication process
introduces in Computer Design. These are:
\begin{itemize}
\item small scale control of placement and connectivity
\item large scal randomness
\item high defect rates
\end{itemize}

This aspects of a DNA-self assembled process leads some architectural implications
when approaching to the Computer Design. Architecture must consist of
a distributed network of small computing and storage nodes, randomly
connected. Data will be travelling on packets that must be routed
using a topology-agnostic strategy, since no particular regularity can
be assumed in such network.

Different works have been presented addressing the problem of topology
agnostic routing, such as~\ref{}. Althougth they show interesting
performances, the requirement for virtual channels resources makes
them not suitable for the particular DNA self assembled we are
focusing on.

Other topology independent algorithms not using virtual channels are
based on turn proibition. Up-Down~\ref{} exploits the creation of a
spanning tree of the topology, then placing bidirectional restrictions
by avoiding a packet to traverse the same link in both up and down
directions. The hierarchical nature of this approach can lead to
uneven traffic distribution, with many packets traversing upper links
(near to the root). The~\ref{} mitigates this issue, making this
approach acceptable in small networks, however the whole restriction
placement is still fixed, depending on the particular root selected,
making these approaches certainly not scalable for the thousands nodes
scenario of DNA self assembled technology.

In~\ref{} authors present SR, a new approach the solves this limitation by
allowing turn restriction to be placed locally, independently from
other restriction. The whole network is partitioned in segments, and a
single bidirectional turn restrictions can be chosen for each
segment in order to guarantee deadlock freedom and connected networks.
This locality independence makes SR scalable for large-sized networks,
but its topology independency still requires the knowledge of the
topology graph in order to find the segments. The small scale control,
random placement and connectivity features discussed in the previous
section makes having such a topology graph impossible in DNA
self assembly, together with any other similar centralized algorithm using
graph description as input.

The contribution presented in this paper is a first effort for a
distribuited segment-based routing in DNA based networks (DiSR).  This
paper is not intended to show discover the "optimal" segment choice, ideally
reachable with a whole graph topology knowledge, but to demonstrate something could
really fit such complex, irregular and large sized topologies.

%------------------------------------------------------------------------------
\section{DiSR main elements}
%------------------------------------------------------------------------------

@@

SEGMENT\_REQUEST: used to search candidates for a segment (not the
first of the subnet, managed separately as a starting segment).

SEGMENT\_CONFIRM: used to create a segment. When received, if the node
status is tvisited with a segID corresponding to the one indicated in
the packet, the node learns to belong to the segment associated to the
id. Further, it should forward this packet to the link where the
original SEGMENT\_REQUEST packet came from, so that all candidate nodes
can learn the segment id they belong to. The following LED shoud be
updated: The state of the node changes from tvisited to visited.  the
link\_tvisited previously set with the segID should be invalidated the
link\_visited variable associated to the incoming confirm should be set
as segID. Note that if the request was flooded across different links,
the link\_visited variable associated to the other paths shouldn’t be
set.  the link\_visited variable associated to the link from which the
request was originated should be set segID

STARTING\_SEGMENT\_REQUEST: the request for the first segment of a
subnet should be differentiated since all nodes (except the starting
one) must forward the packet using a flooding mechanism. This is
necessary since the request packet must reach the starting node, which
is the only node marked as visited. As stated elsewhere, it should
remembered that the first segment of each subnet it's  a loop that
beginning/ending on the starting node.

SEGMENT\_CANCEL: used when the process of searching a segment along a
certain path fails. When a node discovers that the path that includes
the node itself is no more suitable, a SEGMENT\_CANCEL is sent back
across the link from which the SEGMENT\_REQUEST was received.  Possible
reasons for that: The node it’s already candidate for another segment
The node it’s free but can’t find a free link (according the
cyclelinks timeout) The node it’s tvisited for the same request id,
previously forwarded that request along some free direction, but now
it’s receiving a a cancel request and has no more free links

\section{DiSR Execution Model}

@@

\section{Nanoxim and simulation environment}

The simulator Nanoxim is a SystemC tool based on a minimal version of
the Noxim Network-on-Chip simulator. It's not a "faster noxim", but a
drastically reduced noxim the will be able to simulate a random
network on nodes. Current differences from noxim:

removed all routing algorithms that depend on a topology (only XY left
to test meshes) removed flit management, nodes only exchange packets
(e.g. no wormhole)

Future features:

topology agnostic able to test different nano-oriented communication
mechanism

\section{Experiments}
% Absolute results:
% Comparison results:

>>>>

\title{A Distributed Segment-based Approach for Topology-agnostic
Routing in Self-Assembled Nano Networks}

\author{
%  \IEEEauthorblockN{a} \IEEEauthorblockA{Affiliation3\\ email3} \and
%  \IEEEauthorblockN{b} \IEEEauthorblockA{Affiliation2\\ email2}

  }

@@

\begin{abstract}
In this paper we present DiSR, a first effort for a distributed
segment-based approach to routing in a nano-scale topology agnostic
scenario based on DNA self-assembly. The main aim is exploiting the
already-proven qualities of segment-based routing without neither requiring a
topology graph as input, nor needing a centralized algorithm to
discover and configure network paths. Our contribution is not to find
the optimal segments choice, but introducing an approach that can
actually work in such complex, irregular and large sized topologies.
After describing the conceptual elements and the execution model of DiSR,
we use the opensource Nanoxim tool to show some quantitive
comparisons to demonstrate how DiSR still maintains segment-based
qualities while avoiding a centralized tree-based broadcasting and the
usage of expensive virtual channels mechanisms.
\end{abstract}

%------------------------------------------------------------------------------
%\IEEEkeywords{...}

\begin{IEEEkeywords}
Nanotechnology, DNA, Self-assembly, Routing
\end{IEEEkeywords}
%------------------------------------------------------------------------------

\section{Introduction and Motivation}
Exploring long term alternatives to the CMOS technology is gaining
more and more revelance as the scaling trend of such devices keeps
introducing new challenges. Power density, defect tolerance, testing
costs and wire delays are only a few of the many critical aspects
involved~\cite{}. While software parallelism and multicore
approaches~\cite{TODO} are partially mitigating the impact on
performances, it is likely that the growing computing demand will
eventually need even more radical architectural modifications and new
paradigms in order to address the computer design issues of the next
decades.

In the last years, Self-assembled nanoscale architectures~\cite{TOOD}
are emerging as a promising technology due their tera/peta scale of
integration, defect tolerance tolerance and huge potential computing
capabilities~\cite{TODO}. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{TODO}.
The main idea behind this approach is to exploit the physical regularity and
stabiliy of DNA sequences in order to create a scaffold onto which
nano-devices (e.g. nanowires and nanotubes~\cite{TODO}) can be
attached. This can be achieved by designing approapriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold.

A detailed description of the chemical properties involved is far beyond
the scope of this paper (see also~\cite{}), so we will focus on the
three main challenges that this new fabrication process introduces in Computer Design. These are:
limited node resources, large scale randomness and high defect rates.
The first aspect is directly related to the use of complementary DNA
tags in order to place circuit components. In a traditional CMOS
process the complexity is introduced using a photolithography mask, so that larger
circuits require larger masks; instead in self-assembly this
complexity is achieved increasing the number of unique DNA tags, since
more tags means having more control on component placement. Ideally,
by specifying a single different tag for each nano transistor terminal,
we could exactly choose where each component would be placed. But the
number of DNA symbols forming the DNA sequence is limited (sequence
of 4 nucleobases G,A,T,C) and so creating many different tags (of a
given lenght) would mean make them more similar to
each other, increasing the probability of incorrect/partial matching. Then,
limiting the need of many unique tags means putting a limited
complexity in each node. Large scale randomness and defect rates are
the other two fundaments conditions of self-assembled technology:
if, as seen, some complexity can be introduced by designing tag sequences in the
DNA grid of each node, we have no control of where each of these grids
will be placed on the whole network. Further, as a consequence, we cannot guarantee other
typical properties of regular networks, like nodes being connected to 4
neighbors, having a determined orientation and so on.

These aspects of a DNA-self assembly process lead to some important
architectural implications to be addressed when approaching to the
Computer Design: computational model must be founded on a distributed
network of small computing and storage nodes, randomly placed and
interconnected.  Proof-of-concept of such kind of computational model
can be found in~\cite{}, where instructions and data operands travel
on packets and routed using a topology-agnostic strategy that avoids
deadlock, since no particular regularity can be assumed in such
network. Altought we will assume a similar scenario as background, it
is important to underline that the proposed approach to defect
tolerance is more general and not stricly dependent on any of these
architectures.

Different works have been presented addressing the problem of topology
agnostic routing, such as~\cite{}. They show interesting
performances, but the requirement for virtual channels resources 
to avoid deadlock makes them not suitable for the particular DNA self
assembled networks we are focusing on. 
Other topology independent algorithms don't use virtual channels to
achieve deadlock freedom but turn proibitions, like~\cite{}.
Up-Down~\cite{} which exploits the creation of a spanning tree of the
topology, then placing bidirectional restrictions by avoiding a packet
to traverse the same link in both up and down directions. Should be
noticed that the hierarchical nature of this approach can lead to uneven traffic
distribution, with many packets traversing upper links (near to the
root), however this is still acceptable for classical wide area networks
topologies with a limited number of nodes.  The~\cite{} mitigates this
issue, but the whole restriction placement is still prefixed,
striclty depending on the particular root selected, thus making these
approaches not scalable for the TODO(thousands) nodes scenario of DNA
self assembled technology.

In~\cite{} authors present SR, a new approach the solves this limitation by
allowing turn restrictions to be placed locally, independently from
other restrictions. The whole network is partitioned into segments, and a
single bidirectional turn restrictions can be chosen for each
segment in order to guarantee deadlock freedom and connected networks.
This \emph{locality independence} property makes SR scalable for large-sized networks,
but its topology independency still requires the knowledge of the
network topology graph in order to find the segments. The small scale control,
random placement and connectivity features discussed above
make it impossible having such topology graph knowledge, as well as
trying to execute a centralized algorithm that uses the graph description as input.

The contribution presented in this paper is a first effort for a
Distributed Segment-based routing in nanoscale DNA based networks
(DiSR). The presented approach is not intended to discover the "optimal"
segment choice (ideally reachable with the knowledge of the whole
topology graph) but just to demonstrate a concrete model that can fit
into such complex, irregular and large sized topologies.

%------------------------------------------------------------------------------
\section{DiSR main concepts}
%------------------------------------------------------------------------------

@@

\begin{itemize}
\item{SEGMENT\_REQUEST}: used to search candidates for a segment (not the
first of the subnet, managed separately as a starting segment).

\item{SEGMENT\_CONFIRM}: used to create a segment. When received, if the node
status is tvisited with a segID corresponding to the one indicated in
the packet, the node learns to belong to the segment associated to the
id. Further, it should forward this packet to the link where the
original SEGMENT\_REQUEST packet came from, so that all candidate nodes
can learn the segment id they belong to. The following LED shoud be
updated: The state of the node changes from tvisited to visited.  the
link\_tvisited previously set with the segID should be invalidated the
link\_visited variable associated to the incoming confirm should be set
as segID. Note that if the request was flooded across different links,
the link\_visited variable associated to the other paths shouldn’t be
set.  the link\_visited variable associated to the link from which the
request was originated should be set segID

\item{STARTING\_SEGMENT\_REQUEST}: the request for the first segment of a
subnet should be differentiated since all nodes (except the starting
one) must forward the packet using a flooding mechanism. This is
necessary since the request packet must reach the starting node, which
is the only node marked as visited. As stated elsewhere, it should
remembered that the first segment of each subnet it's  a loop that
beginning/ending on the starting node.

\item{SEGMENT\_CANCEL}: used when the process of searching a segment along a
certain path fails. When a node discovers that the path that includes
the node itself is no more suitable, a SEGMENT\_CANCEL is sent back
across the link from which the SEGMENT\_REQUEST was received.  Possible
reasons for that: The node it’s already candidate for another segment
The node it’s free but can’t find a free link (according the
cyclelinks timeout) The node it’s tvisited for the same request id,
previously forwarded that request along some free direction, but now
it’s receiving a a cancel request and has no more free links

\end{itemize}

\section{Draft of DiSR Execution Model}

@@

\section{Simulation environment and Results}

In this section we will test the proposed DiSR approach to demonstrate
its effectiveness from two different perspectives:
\begin{itemize}
\item Against the serial segment based approach SR, to evaluate if 
its qualities are still preserved in the new nano-scale distributed scenario
\item Against other topology agnostic approaches based on spanning
trees and broadcasting, to measure how DiSR performs in discovering
the network structure 
\end{itemize}

\subsection{Nanoxim tool}

In order to quantitatively and qualitatively the proposed approach a
specific simulation environment has been developed, resulting in
the open source and freely available project called Nanoxim~\cite{}.
Nanoxim is a SystemC tool based on a almost rewritten
version of the Noxim Network-on-Chip simulator~\cite{}. While some
complex features have been removed (e.g. wormhole, congestion/topology
aware routing and selection strategies) new features specifically
taylored for nanoscale scenario were introduced, like the ability to simulate a random
network, the implementation of the DiSR to obtain the segment topology
and the support for defective links simulation.

%------------------------------------------------------------------------------
\subsection{Experimental setup}
The following setup parameters have been taken into account when
performing the DiSR simulation:
\begin{itemize}
\item {S}: Size of the network, intended as number of nodes (from X to Y)
\item {P}: probabily of failure, that is the probability that each
link is disconnected or not present 
\item {B}: bootstrap node, the node used from upper (e.g. microscale)
layer to inject the DiSR process

Other optimisation parameters which demostrated to improve the DiSR
results have been fixed to some good values, but are not subject to
further investigation in this initial work, where the focus is not the
optimal set of segments, but just demostrating the working mechaninsm
of the approach. These are:
\begin{itemize}
\item{cycle\_links}: 
\item{bootstrap\_immunity}:
\item{bootstrap\_timeout}:
\end{itemize}

\subsection{Evaluation Metrics and Results}
\label{sec:results}

The following metrics have been adopted:
\begin{itemize}
\item{Node coverage}: this is the fraction of non-defective node elements that
are assigned to a segment. In the ideal case, all the non defective
nodes should be assigned, so this metric is useful to show how some
disconnected node regions can negativaly impact on the whole DiSR
effectiveness.
\item{Broadcast latency}: this measures how the time required to complete the
segment assignment scale for increasing network sizes and defect rates
\item{Connectivity}:
\item{Bootstrap node effect}: this evaluates the impact of the chosen
bootstrap node on the node coverage.
\end{itemize}

Further, to evaluate how DiSR compares against the centralized segment
based algorithm, the following metrics have been adopted:
\begin{itemize}
\item {Average Path Lenght}:
\item {Average Link Weight}:
\item {Number of Segments}:
\end{itemize}

Node (Broadcast) Coverage(vs \%defective links) with different sizes:
30x30, 50x50, 80x80, 100x100, ideal. Bigger networks show similar
trends

Broadcast latency (vs sqrt(N), with different \% defectived
links. TODO: calcolare ordine di grandezza atteso O(N), rete NxN
Effect of bootstrap node

Qualcosa tipo Broadcast Tree properties (sez.  5.4)

\section{Implementation Issues}
\label{sec:implementation}
Althought a completed and detailed floorplan of the node is not the
focus of this work, in this section we want to give a basic idea of
which are the main elements required to support this kind of
architecture and the DiSR approach. There are three main classes of
node elements:
\begin{itemize}
\item Node-specific: components (such ALUs, memories) that are
strictly related to the node functionality and role inside a given
networks: e.g. is that a computation or storage node ?
\item Node communication:  these are elements (such as transceivers,
buffers) required to the node to communicate with its neighbors,
indepentently from node functionality and DiSR implementation
\item DiSR-specific: all the hardware, such as control logic and
configuration, required to implement the proposed DiSR approach.
\end{itemize}

\begin{figure}
  \centering
  \includegraphics[width=0.50\textwidth]{pictures/router_schematic.eps}
  \caption{Schematic of the router. Top-level view (top). Block
  implementing the routing algorithm (middle). Block implementing the
  TODO
  }
  \label{fig:schematic}
\end{figure}

Figure~\ref{fig:schematic} shows the schematic of a router for
a mesh based network topology implementing DiSR . As shown, the top level scheme of the router consists of a
set of ...

The content of a routing algorithm block is more accurately shown in
the central part of Figure~\ref{fig:schematic}. Each routing algorithm
block consists of ...

Let us now look inside the \emph{DiSR} block. A formal description of
DiSRalgorithm has already been given in Figure~\ref{alg:disr}, thus we
will focus on the architectural elements to be implemented.

\end{itemize}
